---
output: html_document
---
# Reproducible Research: Peer Assessment 1

It is now possible to collect a large amount of data about
personal movement using activity monitoring devices such as 
a Fitbit, Nike Fuelband, or Jawbone Up. These type of
devices are part of the "quantified self" movement - a 
group of enthusiasts who take measurements about themselves 
regularly to improve their health, to find patterns in 
their behavior, or because they are tech geeks. But these 
data remain under-utilized both because the raw data are 
hard to obtain and there is a lack of statistical methods 
and software for processing and interpreting the data.

This assignment makes use of data from a personal activity 
monitoring device. This device collects data at 5 minute 
intervals through out the day. The data consists of two 
months of data from an anonymous individual collected 
during the months of October and November, 2012 and include 
the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data

We downloaded the data on 8/7/2014 from URL <- "https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip" and unzipped the file (dated 2/11/2014).  
Here's the code needed to:  
  
1. Load the data.  
2. Process/transform the data into a format suitable for 
our analysis.  

```{r  loading, echo=TRUE}

activ<-read.csv("activity.csv")
activ<-cbind(activ, hourmin=activ[,3])
activ$hourmin<-as.character(activ$hourmin)
activ$hourmin[]<-paste("000",activ$hourmin[],sep="")
activ$hourmin[]<-substr(activ$hourmin[],nchar(activ$hourmin[])-3,nchar(activ$hourmin[]))
activ$hourmin[]<-paste(substr(activ$hourmin[],1,2),substr(activ$hourmin[],3,4),sep=":")

```


## What is the mean total number of steps taken per day?

For this part of the assignment, we ignore the missing 
values in the dataset.  
  
1. We make a histogram of the total number of steps taken 
each day.  

```{r  histogramsteps, echo=TRUE,fig.height=4, fig.width=4}
sumsteps<-tapply(activ$steps,activ$date,sum)
hist(sumsteps, main="Histogram of # of daily steps",xlab="Number of Daily Steps")
```  
  
2. We calculate and report the mean and the median of the total number of steps taken per day.  
```{r  meansteps, echo=TRUE}
meansteps<-mean(sumsteps,na.rm = TRUE)
mediansteps<-median(sumsteps,na.rm = TRUE)
```
The  mean of the total number of daily steps is `r format(meansteps, scientific=FALSE)` and the median is `r format(mediansteps, scientific=FALSE)`.  
  
  
## What is the average daily activity pattern?
  
1. We display a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
2. We then show the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.  
```{r  averagepat, echo=TRUE}
avginterv<-aggregate(activ$steps, by = list(as.factor(activ$hourmin)), FUN = mean, na.rm=T)
plot(avginterv$x, type="l", xlab="Time Interval", ylab="Avg. # of Steps", main="Mean steps per 5-minute interval")
maxsteps<-avginterv$Group.1[avginterv$x==max(avginterv$x)]
maxstepsend<-avginterv$Group.1[which(avginterv$Group.1==maxsteps ) + 1]
```
The 5-minute interval with the maximum numer of steps is `r maxsteps `-`r maxstepsend`.


## Imputing missing values

We should take into account that there are a number of days
/intervals where there are missing values (coded as NA). 
The presence of missing days may introduce bias into some 
calculations or summaries of the data.  
  
1. We calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
2. We decide that we will replace NAs in missing values with the mean daily number of steps per interval.  
3. We then create a new dataset ("activimp") that is equal to the original dataset but with the missing data filled in.  
4. Finally, we make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?  
  
```{r  missing, echo=TRUE}
missval<-sum(!complete.cases(activ))
```
The dataset has `r missval` missing values.  
  
```{r  imputing, echo=TRUE}  
activimp<-activ
for (i in which(sapply(activimp, is.numeric))) {
    activimp[is.na(activimp[, i]), i] <- mean(activimp[, i],  na.rm = TRUE)
}
sumsteps2<-tapply(activimp$steps,activ$date,sum)
hist(sumsteps2, main="Histogram of # of daily steps",xlab="Number of Daily Steps (no NAs)")
meansteps2<-mean(sumsteps2)
mediansteps2<-median(sumsteps2)
```
  
The  mean of the total number of daily steps with imputed data is `r format(meansteps2, scientific=FALSE)` and the median is `r format(mediansteps2, scientific=FALSE)`.   

## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. We use the dataset with the filled-in missing values for this part.  
  
1. We create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  
2. We make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  
```{r  differ, echo=TRUE}
activimp[,5]<-as.POSIXlt(activimp$date)$wday
for (i in 1:nrow(activimp)) {
    if ((activimp[i,5]==0) | (activimp[i,5]==6)) {
        activimp[i,5]<-"weekend"
    }
    else activimp[i,5]<-"weekday"
}
activimp[,5]<-factor(activimp$V5)
par(mfrow = c(2, 1))
activimpd<-activimp[activimp[,5]=="weekday",]
activimpe<-activimp[activimp[,5]=="weekend",]
avgintervd<-aggregate(activimpd$steps, by = list(as.factor(activimpd$hourmin)), FUN = mean, na.rm=T)
plot(avgintervd$x, type="l", xlab="Time Interval", ylab="Avg. # of Steps", main="Mean steps per 5-minute interval on weekdays")
avginterve<-aggregate(activimpe$steps, by = list(as.factor(activimpe$hourmin)), FUN = mean, na.rm=T)
plot(avginterv$x, type="l", xlab="Time Interval", ylab="Avg. # of Steps", main="Mean steps per 5-minute interval on weekends")
```
